---
title: "Lab 06 — Hypothesis Testing from Claims"
author: "EAES 480 — Modern Statistics in Earth & Environmental Science"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
```

# Overview

This lab introduces **hypothesis testing** as an extension of sampling, bootstrap distributions, and uncertainty quantification.

The core idea is simple:

> A **claim** about the population becomes a **null hypothesis**.

You will use the **US-AMS simplified AmeriFlux dataset** again and test claims about population means using:

- one sample,
- a bootstrap estimate of uncertainty,
- a z-statistic,
- and a p-value.

This lab emphasizes **reasoning from claims to tests**, not just mechanics.

---

# Learning goals

By the end of this lab, you should be able to:

- Turn a verbal claim into a null hypothesis
- Choose a variable and define the relevant population parameter
- Draw a sample and estimate a mean
- Use bootstrap resampling to estimate uncertainty in that mean
- Compute a z-statistic and p-value
- Interpret a hypothesis test in plain language
- Connect confidence intervals and hypothesis tests

---

# Data

## Load and inspect

```{r load_data, echo=TRUE, eval=F}
# GOAL: Read the dataset and clean the names.

df <- read_csv("data/us-ams-simple.csv", na = c("-9999")) %>%
  clean_names() %>%
  filter(year_local == 2023)

glimpse(df)
```

## Derive time columns (if needed)

```{r derive_time, echo=TRUE, eval=F}
# GOAL: Create date / month / day_night columns as in earlier labs.

df <- df %>%
  mutate(
    date = as.Date(floor(doy) - 1, origin = paste0(year_local, "-01-01")),
    month = month(date, label = TRUE, abbr = TRUE),
    day_night = if_else(daytime == 1, "Day", "Night")
  )
```

---

# Part I — From claim to hypothesis test

## 1) Choose a variable

Choose **one numeric variable** from the dataset.

Possible examples:
- `swc`
- `ta`
- `fc`
- `gpp`
- `le`

```{r choose_variable, echo=TRUE, eval=T}
target_var <- "gpp"

# CHECK: does this variable exist?
target_var %in% names(df)
```

## 2) Describe the population parameter

Write the population parameter you care about:

> $\mu$ = the true 2023 population mean of `gpp`

Fill in the blank in words:

> In this lab, my population parameter is the true annual mean of **`gpp`** at the US-AMS site.

---

## 3) Make a claim

Write a claim about the population mean for your variable.

Examples:
- “The annual mean soil water content is 0.30.”
- “The annual mean air temperature is 10 °C.”
- “The annual mean latent heat flux is 40 W m$^{-2}$.”

Now turn your claim into hypotheses:

- Null hypothesis: $H_0: \mu = \mu_0$
- Alternative hypothesis: $H_A: \mu \neq \mu_0$

Fill these in:

> My claim is: `The annual mean gross primary productivity is 8.0`

> $H_0$: `There is no significant correlation between ta and gpp.`

> $H_A$: `There is a significant correlation between ta and gpp.`

---

## 4) Draw one sample and compute the sample mean

```{r draw_sample, echo=TRUE, eval=T}
# GOAL: Draw one random sample and compute its mean.

set.seed(123)

n_samp <- 200   # e.g., 200

samp <- df %>%
  filter(!is.na(.data[[target_var]])) %>%
  slice_sample(n = n_samp)

sample_mean <- mean(samp[[target_var]], na.rm = TRUE)

sample_mean
```

**Prompt (2–3 sentences):** How does your sample mean compare to your claim? Is it close, or already quite different?

My sample mean is 7.39 and my claim was 8.0. These two values are relatively close in value within the `gpp` distribution.

---

## 5) Bootstrap the sample mean

```{r bootstrap_means, echo=TRUE, eval=T}
# GOAL: Bootstrap the sample mean to estimate uncertainty.

set.seed(123)

B <- 4000   # e.g., 4000

boot_means <- replicate(
  B,
  mean(sample(samp[[target_var]], size = n_samp, replace = TRUE), na.rm = TRUE)
)

head(boot_means)
```

## 6) Estimate uncertainty

```{r bootstrap_summary, echo=TRUE, eval=T}
# GOAL: Compute the bootstrap SE and 95% CI.

boot_se <- sd(boot_means)
boot_ci <- quantile(boot_means, probs = c(0.025, 0.975))
tibble(
  sample_mean = sample_mean,
  boot_se = boot_se,
  ci_lower = boot_ci[1],
  ci_upper = boot_ci[2]
)
```

**Prompt (2–3 sentences):** What does the bootstrap SE tell you? What does the 95% CI tell you?

The bootstrap SE is relatively small, meaning the estimate of the population mean was more precise. 
The 95% CI implies that through repetition of the study, there's a 95% confidence intervals would have the true population value. 
---

## 7) Convert the claim into a formal test

Define the null value from your claim.

```{r hypothesis_test, echo=TRUE, eval=T}
# GOAL: Compute z and p-value for your claim.

mu0 <- 8.0   # numeric claim value

z_stat <- (sample_mean - mu0) / boot_se

p_value <- 2 * pnorm(-abs(z_stat))

tibble(
  sample_mean = sample_mean,
  mu0 = mu0,
  boot_se = boot_se,
  z_stat = z_stat,
  p_value = p_value
)
```
---

## 8) Interpret the result

Use your p-value and CI to answer:

- Does the 95% CI include the null value?
- Is the p-value small or large?
- Would you reject the null hypothesis at $\alpha = 0.05$?

Fill in:

> The null value `8.0` is inside my 95% CI.

> My p-value is `0.446`, which is large.

> Therefore, I would not reject $H_0$ at the 0.05 level.

**Prompt (4–6 sentences):** Explain your decision in plain language. Do the data make the claim look plausible or surprising? Be careful: “fail to reject” does not mean “prove true.”

Since the null value falls within the 95% CI and the p-value is much higher than 0.05, the data provides no strong evidence against the null hypothesis. The results would suggest the data is quite consistent with the null hypothesis. This does not prove that the null hypothesis is "true", but rather that the data is not demonstrating strong differences, providing reason to claim the null hypothesis is reasonable. 
---

## 9) Visualize the hypothesis test

```{r plot_test, echo=TRUE, eval=T}
# GOAL: Plot the bootstrap distribution, sample mean, CI, and null value.

ggplot(tibble(boot_mean = boot_means), aes(x = boot_mean)) +
  geom_histogram(bins = 25, alpha = 0.8) +
  geom_vline(xintercept = sample_mean, linetype = "dashed", linewidth = 1.1) +
  geom_vline(xintercept = boot_ci, color = "red", linewidth = 1.1) +
  geom_vline(xintercept = mu0, color = "blue", linewidth = 1.1) +
  theme_classic(base_size = 18) +
  labs(
    x = paste0("Bootstrap mean of ", target_var),
    y = "Count",
    title = "Bootstrap distribution and null claim",
    subtitle = "Dashed = sample mean; Red = 95% CI; Blue = null value"
  )
```

**Prompt (2–3 sentences):** Does the plot agree with your p-value-based conclusion? Explain.

The plot agrees with my p-value-based conclusion. The confidence intervals align with the the failure to reject the null hypothesis. 
---

## 10) Try a second claim (short extension)

Keep the same variable, but test a **different null value**.

```{r second_claim, echo=TRUE, eval=T}
# GOAL: Test a second claim using the same sample and bootstrap SE.

mu0_alt <- 6.0

z_stat_alt <- (sample_mean - mu0_alt) / boot_se
p_value_alt <- 2 * pnorm(-abs(z_stat_alt))

tibble(
  mu0_alt = mu0_alt,
  z_stat_alt = z_stat_alt,
  p_value_alt = p_value_alt
)
```

**Prompt (3–4 sentences):** How did changing the null value change the conclusion? What does this tell you about the relationship between the null hypothesis and the observed estimate?

Decreasing the null value to below the population mean decreased the p-value. However it is still failing to reject the null hypothesis (>0.05). This provides evidence that the null hypothesis is not fixed, but helps create a benchmark for measuring evidence. More specifically, when the null hypothesis is closer to the observed mean/estimate, the harder it is to reject. 
---

# Part II — Comparing group means with t-tests

In Part I, you tested a claim about **one population mean**.

Now you will extend the same logic to compare **two group means**.

The key idea is:

> A two-sample t-test asks whether the difference between two group means is large relative to the uncertainty in that difference.

You will first work with a built-in grouping variable (**day vs night**), then choose a second grouping variable of your own.

---

## 11) Create a day/night grouping variable

If you already created `day_night` earlier, you can reuse it. Otherwise, create it here.

```{r make_daynight, echo=TRUE, eval=FALSE}
# GOAL: create a day/night grouping variable from `daytime`.

df <- df %>%
  mutate(
    day_night = if_else(daytime == 1, "Day", "Night")
  ) %>% 
  filter(!is.na(daytime))

df %>% count(day_night)
```

---

## 12) Choose a response variable for the t-test

Pick one numeric response variable to compare between groups.

Possible examples:
- `ta`
- `swc`
- `fc`
- `gpp`
- `le`

```{r choose_response_ttest, echo=TRUE, eval=FALSE}
response_var <- "ta"

# CHECK: does this variable exist?
response_var %in% names(df)
```

---

## 13) Visualize the response by day/night

Create a plot that shows how your response variable differs between Day and Night.

You may use:
- `geom_boxplot()`
- `geom_violin()`
- `geom_jitter()`
- or a combination

```{r plot_daynight, echo=TRUE, eval=T}
# GOAL: visualize the response by day/night group.

df %>%
  filter(!is.na(.data[[response_var]])) %>%
  ggplot(aes(x = day_night, y = .data[[response_var]], fill = day_night)) +
  geom_boxplot(alpha = 0.8) +
  theme_classic(base_size = 18) +
  labs(
    x = NULL,
    y = response_var,
    title = "Response by day vs night"
  )
```

**Prompt (2–3 sentences):** Before running the test, what do you predict?  
Do the groups look different in center? In spread?

I believe that the groups will look different in spread and centers. The day will be more drastic compared to the night data, since the hottest air temperatures/most fluctuation occurs during the day. The night time will have a smaller spread and the center lower.
---

## 14) Summarize the two groups

Compute the group means, SDs, and sample sizes.

```{r summarize_daynight, echo=TRUE, eval=T}
# GOAL: summarize the response variable by day/night group.

group_summary <- df %>%
  filter(!is.na(.data[[response_var]])) %>%
  group_by(day_night) %>%
  summarise(
    mean_response = mean(.data[[response_var]], na.rm = TRUE),
    sd_response = sd(.data[[response_var]], na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

group_summary
```

---

## 15) State the hypotheses for the day/night comparison

Write hypotheses in words and symbols.

Use this structure:

- Null: the two group means are equal
- Alternative: the two group means are different

Fill in:

> My response variable is `ta`

> $H_0$: `There is no significant relationship between ta and the time of day.`

> $H_A$: `There is a significant relationship between ta and the time of day.`

---

## 16) Run a two-sample t-test

Use `t.test()` to compare the means of your response variable between Day and Night.

```{r ttest_daynight, echo=TRUE, eval=T}
# GOAL: run a two-sample t-test by formula.

t.test(
  response_var ~ day_night,
  data = df
)
```

**Prompt (4–6 sentences):** Report the p-value and confidence interval.  
Would you reject the null hypothesis at $\alpha = 0.05$?  
What does the result imply about day vs night differences in your chosen variable?

> *Write your answer here.*

---

## 17) Connect the t-test to a confidence interval

Extract the confidence interval and interpret it as a difference in means.

```{r ttest_store, echo=TRUE, eval=T}
# GOAL: store the t-test result in an object.

tt_daynight <- t.test(
  ___ ~ ___,
  data = ___
)

tt_daynight$conf.int
```

**Prompt (2–3 sentences):** Does the confidence interval include 0?  
How does that connect to the hypothesis test decision?

> *Write your answer here.*

---

## 18) Bootstrap view of the same comparison (conceptual extension)

Create bootstrap distributions of the **group means** for Day and Night separately.

```{r bootstrap_group_means, echo=TRUE, eval=T}
set.seed(123)

B <- 2000   # e.g., 2000

x_day <- df %>%
  filter(day_night == "Day") %>%
  pull(___) %>%
  na.omit()

x_night <- df %>%
  filter(day_night == "Night") %>%
  pull(___) %>%
  na.omit()

boot_day <- replicate(
  B,
  mean(sample(x_day, size = length(x_day), replace = TRUE))
)

boot_night <- replicate(
  B,
  mean(sample(x_night, size = length(x_night), replace = TRUE))
)

boot_groups <- bind_rows(
  tibble(group = "Day", mean_est = boot_day),
  tibble(group = "Night", mean_est = boot_night)
)
```

```{r plot_boot_groups, echo=TRUE, eval=FALSE}
# GOAL: visualize the bootstrap distributions by group.

ggplot(boot_groups, aes(x = mean_est, fill = group)) +
  geom_histogram(alpha = 0.45, bins = 25, position = "identity") +
  theme_classic(base_size = 18) +
  labs(
    x = paste0("Bootstrap mean of ", response_var),
    y = "Count",
    title = "Bootstrap distributions of group means"
  )
```

**Prompt (3–4 sentences):** How does this bootstrap visualization help you understand the t-test result?  
Do the plausible group means strongly overlap, or are they clearly separated?

> *Write your answer here.*

---

## 19) Choose a second grouping variable of your own

Now define **one additional grouping variable** and repeat the t-test workflow.

Possible ideas:
- month grouped into two seasons (e.g., winter vs summer)
- a high/low split of another continuous variable
- a wet/dry split of SWC
- a daytime subset grouped by month category

You must:
- define the grouping variable clearly
- explain why the grouping is scientifically meaningful

```{r choose_second_group, echo=TRUE, eval=FALSE}
# GOAL: create your own two-level grouping variable.

df <- df %>%
  mutate(
    my_group = ___
  )

df %>% count(my_group)
```

**Prompt (2–3 sentences):** Why did you choose this grouping variable? What scientific contrast are you testing?

> *Write your answer here.*

---

## 20) Visualize and test your second grouping variable

Repeat the workflow:
1. visualize
2. summarize
3. state hypotheses
4. run `t.test()`
5. interpret

```{r visualize_second_group, echo=TRUE, eval=FALSE}
# GOAL: visualize your response by your chosen grouping variable.

___
```

```{r summarize_second_group, echo=TRUE, eval=FALSE}
# GOAL: summarize your response variable by your chosen grouping variable.

___
```

```{r ttest_second_group, echo=TRUE, eval=FALSE}
# GOAL: run a two-sample t-test for your chosen grouping variable.

___
```

**Prompt (5–6 sentences):** Compare the day/night t-test to your second grouping-variable test.  
Which contrast was stronger? Which was more scientifically meaningful?  
Did the two tests produce similar or different conclusions?

> *Write your answer here.*

---

## 21) Final reflection

Answer in **4–6 sentences**:

- How is a two-sample t-test similar to the one-mean hypothesis test from Part I?
- What changes when we move from “testing one mean” to “testing a difference in means”?
- How did the bootstrap group-mean plot help your understanding?

> *Write your answer here.*

---

# Submission

- Knit your `.Rmd` to HTML.
- Commit and push both the `.Rmd` and knitted `.html` to your GitHub repo.
- Submit the GitHub repository link on Canvas.